---
title: "Figure S2: Extended analyses of cells programs and intercellular signaling"
author: "Lin-Xiong Ye (445233812@qq.com)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)
```

# Overview

This notebook contains all analyses for **Figure S2** (Supplementary Figure 2).

**Figure panels:**
- **S2A-B**: Volcano plots for SMC2 and SMC3
- **S2C-F**: Trajectory analysis details
- **S2E**: MYH11 and ACTA2 expression
- **S2F**: CLU and LGALS1 expression
- **S2G-I**: Endothelial cell analysis
- **S2J**: CellChat summary statistics
- **S2K**: Ligand/receptor expression
- **S2L**: Ligand/receptor expression heatmap
- **S2M**: SMC2/3 to EC signaling flow

---

# Load Libraries

```{r load-libraries}
library(Seurat)
library(ggplot2)
library(dplyr)
library(SCP)
library(tidyverse)
library(CellChat)
library(ComplexHeatmap)
library(circlize)
library(EnhancedVolcano)
library(gridExtra)

options(future.globals.maxSize = 14000000 * 1024^2)
```

---

# Load Data

```{r load-data}
cat("=== Loading Data ===\n")

# Load preprocessed data
seurat_obj <- readRDS('../seurat_object.rds')
result_cellchat <- readRDS('../cellchat_results.rds')

cat("Data loaded successfully!\n")
```

---

#-------------------------------------------------------------------------------
# Figure S2A-B: Volcano plots for SMC2 and SMC3
#-------------------------------------------------------------------------------
cat("\n--- Figure S2A-B: SMC subtype volcano plots ---\n")

smc2 = subset(seurat_obj, subtype %in% c('SMC2'))
smc3 = subset(seurat_obj, subtype %in% c('SMC3'))

smc2 = RunDEtest(srt = smc2, group.by = "exp",
                 fc.threshold = 1, only.pos = F)
de_filter = smc2@tools[["DEtest_custom"]][["AllMarkers_wilcox"]]
de_top <- de_filter %>% group_by(group1) %>% slice_head(n = 5) 


de_filter$threshold <- as.factor(ifelse(de_filter$avg_log2FC >= 1 & de_filter$p_val_adj < 0.05, 
                                        'Up', ifelse(de_filter$avg_log2FC <= -1 & de_filter$p_val_adj < 0.05,
                                                     'Down', 'Not')))
de_filter <- data.frame(de_filter)

top_up_genes <- de_filter %>% filter(threshold == 'Up') %>% arrange(p_val_adj) %>% head(5)
top_down_genes <- de_filter %>% filter(threshold == 'Down') %>% arrange(p_val_adj) %>% head(5)

top_genes <- bind_rows(top_up_genes, top_down_genes)

library(EnhancedVolcano)
p_s2a = EnhancedVolcano(de_filter, x = 'avg_log2FC',
                        y = 'p_val_adj', drawConnectors = F,
                        lab = de_filter$gene, xlim = c(-3, 3), 
                        FCcutoff = 0.5, legendPosition = 'None', 
                        selectLab =  top_genes$gene,
                        subtitle = NULL, title = paste("SMC2"),
                        widthConnectors = 0.2,
                        lengthConnectors = unit(0.0, 'npc'), labSize = 2.0,pointSize = 1)



ggsave("figures/figureS2/FigS2A_SMC2_volcano.pdf", p_s2a, width = 10, height = 8)

ggsave("figures/figureS2/FigS2B_SMC3_volcano.pdf", p_s2b, width = 10, height = 8)
cat("Saved: FigS2A-B volcano plots\n")

#-------------------------------------------------------------------------------
# Figure S2C-F: Trajectory analysis details
#-------------------------------------------------------------------------------
cat("\n--- Figure S2C-F: Trajectory details ---\n")
cat("Note: Requires Monocle3 trajectory pre-processing\n")
cat("Refer to original scripts for detailed trajectory code\n")

#-------------------------------------------------------------------------------
# Figure S2E: MYH11 and ACTA2 expression across SMC subtypes
#-------------------------------------------------------------------------------
cat("\n--- Figure S2E: Contractile markers ---\n")

p_s2e <- FeatureStatPlot(smc.mergeTEN,
                         stat.by = c('MYH11', 'ACTA2'),
                         group.by = "subtype",
                         comparisons = FALSE,
                         plot_type = "violin")

ggsave("figures/figureS2/FigS2E_contractile_markers.pdf", p_s2e, width = 8, height = 4)
cat("Saved: FigS2E_contractile_markers.pdf\n")


#-------------------------------------------------------------------------------
# Figure S2F: CLU and LGALS1 expression
#-------------------------------------------------------------------------------

p = DynamicPlot(
  srt = smc.mergeTEN, lineages = c("Monocle3_Pseudotime"), group.by = "subtype",
  features = c("CLU", "LGALS1"),
  compare_lineages = TRUE, compare_features = FALSE
)

ggsave(file = 'figures/figureS2/genes.pdf',p,width = 6,height = 3)


#-------------------------------------------------------------------------------
# Figure S2G-I: Endothelial cell analysis
#-------------------------------------------------------------------------------
cat("\n--- Figure S2G-I: Endothelial cell analysis ---\n")

# Subset endothelial cells
ec <- subset(seurat_obj, subtype %in% c("EC1", "EC2", "EC3"))

# UMAP
p_s2g <- DimPlot(ec,
                 reduction = "umap",
                 group.by = "subtype",
                 label = TRUE)

ggsave("figures/figureS2/FigS2G_EC_UMAP.pdf", p_s2g, width = 7, height = 6)

# Proportions
p_s2h <- CellStatPlot(ec,
                      stat.by = "subtype",
                      group.by = "exp",
                      plot_type = "trend",
                      label = FALSE)

ggsave("figures/figureS2/FigS2H_EC_proportions.pdf", p_s2h, width = 4, height = 4)

# GO enrichment
ec <- RunEnrichment(
  srt = ec,
  group_by = "subtype",
  db = "GO_BP",
  species = "Homo_sapiens"
)

p_s2i <- EnrichmentPlot(
  srt = ec,
  group_by = "subtype",
  db = "GO_BP",
  plot_type = "comparison"
)

ggsave("figures/figureS2/FigS2I_EC_GO.pdf", p_s2i, width = 8, height = 7)
cat("Saved: FigS2G-I EC analysis\n")

#-------------------------------------------------------------------------------
# Figure S2J: CellChat summary statistics
#-------------------------------------------------------------------------------
cat("\n--- Figure S2J: CellChat summary ---\n")
cat("Note: Requires detailed CellChat analysis\n")

#-------------------------------------------------------------------------------
# Figure S2K: Violin plots of ligand/receptor genes
#-------------------------------------------------------------------------------
cat("\n--- Figure S2K: Ligand/receptor expression ---\n")

ligand_receptor_genes <- c('SEMA4C', 'JAG1', 'VEGFA', 'APP',
                           'NOTCH1', 'CD74', 'CDH5', 'FLT1')

p_s2k <- FeatureStatPlot(seurat_obj,
                         stat.by = ligand_receptor_genes,
                         legend.position = "top",
                         legend.direction = "horizontal",
                         split.by = 'exp',
                         group.by = "subtype",
                         bg.by = "subtype",
                         stack = TRUE,
                         comparisons = TRUE)

ggsave("figures/figureS2/FigS2K_ligand_receptor.pdf", p_s2k, width = 8, height = 6)
cat("Saved: FigS2K_ligand_receptor.pdf\n")

#-------------------------------------------------------------------------------
# Figure S2L: Heatmap of ligand/receptor expression changes (log2FC + P-values)
#-------------------------------------------------------------------------------
cat("\n--- Figure S2L: Ligand/receptor expression heatmap ---\n")

# Define pathway genes (ligands and receptors)
pathway_genes <- c("SEMA4C", "PLXNB2",
                   "VEGFA", "FLT1",
                   "JAG1", "NOTCH1",
                   "APP", "CD74",
                   "CDH5", "ANGPTL1")
available_genes <- pathway_genes[pathway_genes %in% rownames(seurat_obj)]

# Calculate statistics for each gene and cell type
expr_data <- FetchData(seurat_obj, vars = c(available_genes, "subtype", "exp"))
cell_types <- sort(unique(expr_data$subtype))

stats_list <- list()

for (gene in available_genes) {
  for (ct in cell_types) {
    ct_data <- expr_data %>% filter(subtype == ct)

    normal_expr <- ct_data %>% filter(exp == "Adjacent normal") %>% pull(!!gene)
    ad_expr     <- ct_data %>% filter(exp == "Dissection tissue") %>% pull(!!gene)

    # Remove NA
    normal_expr <- normal_expr[!is.na(normal_expr)]
    ad_expr     <- ad_expr[!is.na(ad_expr)]

    # Wilcoxon test
    pval <- NA
    if (length(normal_expr) > 3 && length(ad_expr) > 3) {
      pval <- suppressWarnings(wilcox.test(ad_expr, normal_expr)$p.value)
    }

    mn_norm <- mean(normal_expr, na.rm = TRUE)
    mn_ad   <- mean(ad_expr, na.rm = TRUE)

    stats_list[[length(stats_list) + 1]] <- data.frame(
      Gene = gene,
      CellType = ct,
      Mean_Normal = mn_norm,
      Mean_AD = mn_ad,
      Log2FC = log2((mn_ad + 0.01) / (mn_norm + 0.01)),
      Pvalue = pval,
      stringsAsFactors = FALSE
    )
  }
}

stats_table <- bind_rows(stats_list)

# Create matrices for heatmap
log2fc_matrix <- stats_table %>%
  select(Gene, CellType, Log2FC) %>%
  pivot_wider(names_from = CellType, values_from = Log2FC) %>%
  column_to_rownames("Gene") %>%
  as.matrix()

pval_matrix <- stats_table %>%
  select(Gene, CellType, Pvalue) %>%
  pivot_wider(names_from = CellType, values_from = Pvalue) %>%
  column_to_rownames("Gene") %>%
  as.matrix()

# Order columns (remove non-existent cell types)
col_order <- c("Endothelial cells", "Fibroblasts",
               "SMC1", "SMC2", "SMC3", "SMC4", "SMC5")
col_order <- col_order[col_order %in% colnames(log2fc_matrix)]

log2fc_matrix <- log2fc_matrix[available_genes, col_order, drop = FALSE]
pval_matrix   <- pval_matrix[available_genes, col_order, drop = FALSE]

# Format P-values
fmt_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p < 0.001) return(sprintf("%.1e", p))
  if (p < 0.01)  return(sprintf("%.3f", p))
  sprintf("%.2f", p)
}

# Cell annotation function
cell_fun_l2fc_p <- function(j, i, x, y, width, height, fill) {
  l2fc <- log2fc_matrix[i, j]
  pval <- pval_matrix[i, j]

  if (is.na(l2fc)) return(NULL)

  lab1 <- sprintf("Log2FC=%.2f", l2fc)
  lab2 <- if (!is.na(pval)) paste0("p=", fmt_p(pval)) else ""

  grid.text(lab1, x, y + unit(1.6, "mm"),
            gp = gpar(fontsize = 6, col = "black"))
  if (lab2 != "") {
    grid.text(lab2, x, y - unit(1.6, "mm"),
              gp = gpar(fontsize = 6, col = "black"))
  }
}

# Create heatmap
col_fun <- colorRamp2(c(-1.5, 0, 1.5), c("steelblue", "white", "firebrick"))

ht_s2l <- Heatmap(
  log2fc_matrix,
  name = "Log2FC",
  col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_names_side = "left",
  column_names_rot = 45,
  cell_fun = cell_fun_l2fc_p,
  row_title = "Genes (Ligands & Receptors)",
  column_title = "Log2FC (AD vs Normal)",
  row_names_gp = gpar(fontface = "italic"),
  heatmap_legend_param = list(
    title = "Log2FC\n(AD/Normal)",
    at = c(-1.5, -0.5, 0, 0.5, 1.5)
  )
)

pdf("figures/figureS2/FigS2L_LR_expression_heatmap.pdf", width = 8, height = 5)
draw(ht_s2l)
dev.off()

cat("Saved: FigS2L_LR_expression_heatmap.pdf\n")

#-------------------------------------------------------------------------------
# Figure S2M: Relative information flow from SMC2/SMC3 to EC (AD vs Normal)
#-------------------------------------------------------------------------------
cat("\n--- Figure S2M: SMC2/3 to EC signaling flow ---\n")

# Use merged CellChat object
object.list <- list(Normal = result_cellchat$cellchat_control,
                    AD = result_cellchat$cellchat_test)
cellchat <- mergeCellChat(object.list, add.names = names(object.list))

# Get cell types
cell_types <- levels(cellchat@idents[[1]])

# Create signaling flow comparison for SMC2 and SMC3 to EC
pdf("figures/figureS2/FigS2M_SMC23_EC_signaling_flow.pdf", width = 12, height = 10)

gg_smc_ec <- list()

for(smc in c("SMC2", "SMC3")) {
  if(smc %in% cell_types & "Endothelial cells" %in% cell_types) {
    gg_smc_ec[[smc]] <- tryCatch({
      rankNet(cellchat,
              mode = "comparison",
              measure = "weight",
              color.use = c('steelblue', 'firebrick'),
              sources.use = smc,
              targets.use = "Endothelial cells",
              stacked = TRUE,
              do.stat = TRUE,
              title = paste(smc, "→ Endothelial cells\n(Normal vs AD)"))
    }, error = function(e) {
      message(paste("Error for", smc, ":", e$message))
      NULL
    })
  }
}

# Remove NULL entries
gg_smc_ec <- gg_smc_ec[!sapply(gg_smc_ec, is.null)]

if(length(gg_smc_ec) > 0) {
  grid.arrange(grobs = gg_smc_ec, ncol = 2)
}

dev.off()

cat("Saved: FigS2M_SMC23_EC_signaling_flow.pdf\n")

cat("\n=== FIGURE S2 COMPLETE ===\n")

---

# Session Info

```{r session-info}
sessionInfo()
```
