---
title: "Figure S1: Supplementary marker, clinical association, and myeloid enrichment analyses"
author: "Lin-Xiong Ye (445233812@qq.com)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)
```

# Overview

This notebook contains all analyses for **Figure S1** (Supplementary Figure 1).

**Figure panels:**
- **S1A**: Canonical markers for cell type annotation
- **S1B**: TOP10 genes vs clinical features
- **S1C**: GO-BP enrichment across myeloid subtypes

---

# Load Libraries

```{r load-libraries}
library(Seurat)
library(ggplot2)
library(dplyr)
library(SCP)
library(tidyverse)
library(ggpubr)
library(patchwork)

options(future.globals.maxSize = 14000000 * 1024^2)
```

---

# Load Data

```{r load-data}
cat("=== Loading Data ===\n")

# Load preprocessed data
seurat_obj <- readRDS('../seurat_object.rds')
myeloid_obj <- readRDS('../myeloid_subset.rds')

cat("Data loaded successfully!\n")
```

---

#-------------------------------------------------------------------------------
# Figure S1A: Canonical markers for annotation
#-------------------------------------------------------------------------------
cat("\n--- Figure S1A: Canonical markers ---\n")
# 1) Myeloid
myeloid_markers <- c("CD68", "CD14", "MARCO", "VCAN", "S100A8")

# 2) Lymphocytes
lymphocyte_markers <- c("PTPRC", "CD3D", "MS4A1", "CD79A", "NKG7")

# 3) Fibroblasts
fibroblast_markers <- c("DCN", "LUM", "PDGFRA", "FAP") 

# 4) Endothelial
endothelial_markers <- c("PECAM1", "VWF", "CDH5")

# 5) SMC
smc_markers <- c("ACTA2", "TAGLN", "MYH11")

marker_list <- list(
  "Myeloid"     = myeloid_markers,
  "Lymphocytes" = lymphocyte_markers,
  "Fibroblasts" = fibroblast_markers,
  "Endothelial" = endothelial_markers,
  "SMC"         = smc_markers
)

seen <- character(0)
marker_list <- lapply(marker_list, function(x) {
  x <- x[x %in% rownames(seurat_obj)] 
  x_unique <- setdiff(x, seen)
  seen <<- c(seen, x_unique)
  return(x_unique)
})
marker_list <- marker_list[sapply(marker_list, length) > 0]

print(marker_list)

library(ggplot2)

p_marker_dot <- DotPlot(
  seurat_obj,          
  features = marker_list, 
  group.by = "main_type",    
  cols = c("lightgrey", "firebrick"), 
  dot.scale = 8
) +
  RotatedAxis() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.grid.major = element_line(colour = "grey90", linetype = "dashed"),
    
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "italic"), 
    axis.text.y = element_text(size = 10),
    
    strip.text.x = element_blank(),      
    strip.background = element_blank(),  
    
    legend.position = "right"
  ) +
  labs(
    title = NULL, 
    x = NULL,     
    y = NULL     
  )

print(p_marker_dot)

ggsave("figures/figureS1/FigS1A_Canonical_markers.pdf",
       p_marker_dot,
       width = 12,
       height = 6)

#-------------------------------------------------------------------------------
# Figure S1B: TOP10 genes vs clinical features
#-------------------------------------------------------------------------------
cat("\n--- Figure S1B: TOP10 genes vs clinical variables ---\n")

# Get TOP10 upregulated genes in thrombus-positive samples
top10_up <- de_results_smc %>%
  filter(group1 == "Yes") %>%
  filter(avg_log2FC > 0 & p_val_adj < 0.05) %>%
  arrange(desc(avg_log2FC)) %>%
  head(10)

cat("\nTOP10 upregulated genes:\n")
print(top10_up[, c("gene", "avg_log2FC", "p_val_adj")])

top10_genes <- top10_up$gene

markers_smc_ad
top_genes

smc_ad_subset$Involvement_of_Valves <- factor(smc_ad_subset$Involvement_of_Valves,
                                         levels = c("No", "Yes"))
smc_ad_subset$Involvement_of_Coronary_and_Important_Vessels <- factor(smc_ad_subset$Involvement_of_Coronary_and_Important_Vessels,
                                              levels = c("No", "Yes"))
smc_ad_subset$Pericardial_Effusion <- factor(smc_ad_subset$Pericardial_Effusion,
                                              levels = c("No", "Yes"))

# Clinical variables to test
clinical_groups <- c("Presence_of_Thrombus",
                     "Involvement_of_Valves",
                     "Involvement_of_Coronary_and_Important_Vessels",
                     "Pericardial_Effusion")

# Create combined plot
plot_list <- list()

for (i in seq_along(clinical_groups)) {
  group_col <- clinical_groups[i]

  p <- FeatureStatPlot(
    srt = smc_ad_subset,
    stat.by = top_genes,
    group.by = group_col,
    sig_label = "p.format",
    bg_alpha = 0,
    multiplegroup_comparisons = TRUE,
    stack = TRUE,
    plot_type = "violin"
  ) +
    theme(
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA)
    )

  if (i > 1) {
    p <- p + theme(
      axis.text.y = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks.y = element_blank()
    )
  }

  plot_list[[i]] <- p
}

combined_plot <- wrap_plots(plot_list, nrow = 1)

ggsave("figures/figureS1/FigS1B_TOP10_clinical.pdf",
       combined_plot,
       width = 14,
       height = 8)

cat("Saved: FigS1B_TOP10_clinical.pdf\n")



#-------------------------------------------------------------------------------
# Figure S1C: GO-BP enrichment across myeloid subtypes
#-------------------------------------------------------------------------------
cat("\n--- Figure S1C: Myeloid GO enrichment ---\n")

# Run enrichment for myeloid cells

myeloid_obj = RunDEtest(srt = myeloid_obj, group.by = "celltype",
                   fc.threshold = 1.5, only.pos = F)

myeloid_obj <- RunEnrichment(
  srt = myeloid_obj, group.by = "celltype", db =  c('GO_BP'), species = "Homo_sapiens",
  DE_threshold = "avg_log2FC > log2(1.5) & p_val_adj < 0.05"
)

p_s1c = EnrichmentPlot(srt = myeloid_obj, group.by = "celltype", plot_type = "comparison")

ggsave("figures/figureS1/FigS1C_myeloid_GO.pdf", p_s1c, width = 12, height = 10)
cat("Saved: FigS1C_myeloid_GO.pdf\n")

cat("\n=== FIGURE S1 COMPLETE ===\n")

---

# Session Info

```{r session-info}
sessionInfo()
```
