---
title: "Figure 1: Single-cell RNA sequencing reveals cellular heterogeneity in thrombus-related Aortic Dissection"
author: "Lin-Xiong Ye (445233812@qq.com)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)
```

# Overview

This notebook contains all analyses for **Figure 1**, which demonstrates cellular heterogeneity in thrombus-related aortic dissection using single-cell RNA sequencing.

**Figure panels:**
- **1A**: UMAP visualization of all cell types
- **1B**: Top marker genes heatmap
- **1C**: Cell composition by thrombus status
- **1D**: Volcano plot of DEGs
- **1E**: GSEA analysis
- **1F**: ANGPTL4 expression
- **1G**: Myeloid UMAP
- **1H**: Myeloid proportions

---

# Load Libraries

```{r load-libraries}
library(Seurat)
library(ggplot2)
library(dplyr)
library(harmony)
library(SCP)
library(tidyverse)
library(ggpubr)
library(patchwork)
library(EnhancedVolcano)
library(rstatix)

# Set options
options(future.globals.maxSize = 14000000 * 1024^2)
```

---

# Load Data

```{r load-data}
cat("=== Loading Data ===\n")

# Load preprocessed data
# Note: Replace these file paths with your actual data file locations
seurat_obj <- readRDS('../seurat_object.rds')
myeloid_obj <- readRDS('../myeloid_subset.rds')

cat("Data loaded successfully!\n")
cat("Total cells:", ncol(seurat_obj), "\n")
cat("Cell types:\n")
print(table(seurat_obj$subtype))
```

---

# Figure 1A: UMAP of All Cell Types

```{r fig1a-umap}
cat("\n--- Figure 1A: UMAP by cell types ---\n")

p1a <- CellDimPlot(seurat_obj,
                   reduction = "umap",
                   group.by = "subtype",
                   label = TRUE,
                   raster = TRUE,
                   label.size = 3,
                   pt.size = 0.1) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave("Fig1A_UMAP_celltypes.pdf", p1a, width = 8, height = 7)
cat("Saved: Fig1A_UMAP_celltypes.pdf\n")

# Export source data
fig1a_data <- data.frame(
  UMAP_1 = seurat_obj@reductions$umap@cell.embeddings[, 1],
  UMAP_2 = seurat_obj@reductions$umap@cell.embeddings[, 2],
  Cell_Type = seurat_obj$subtype
)
write.csv(fig1a_data, "Fig1A_UMAP_coordinates.csv", row.names = FALSE)
cat("Exported: Fig1A source data (", nrow(fig1a_data), "cells)\n")

print(p1a)
```

---

# Figure 1B: Top Marker Genes Heatmap

```{r fig1b-markers}
cat("\n--- Figure 1B: TOP5 marker genes for each cell type ---\n")

# Run differential expression test
seurat_obj <- RunDEtest(srt = seurat_obj,
                          group_by = "subtype",
                          fc.threshold = 1,
                          only.pos = TRUE)

# Extract TOP5 markers per cell type
de_filter <- seurat_obj@tools[["DEtest_subtype"]][["AllMarkers_wilcox"]]
de_top5 <- de_filter %>%
  group_by(group1) %>%
  slice_head(n = 5)

top5_genes <- unique(de_top5$gene)

cat("\nTOP5 markers per cell type:\n")
print(de_top5[, c("group1", "gene", "avg_log2FC", "p_val_adj")])

# Create heatmap
p1b <- GroupHeatmap(
  srt = seurat_obj,
  features = top5_genes,
  group.by = "subtype",
  heatmap_palette = "YlOrRd",
  cell_annotation_palette = c("Dark2", "Paired"),
  dot_size = unit(4.8, "mm"),
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_side = "left",
  add_dot = TRUE,
  add_reticle = TRUE
)

ggsave("Fig1B_TOP5_markers_heatmap.pdf",
       p1b$plot,
       width = 6,
       height = 6)

cat("Saved: Fig1B_TOP5_markers_heatmap.pdf\n")

# Export source data
fig1b_data <- de_top5 %>%
  select(Cell_Type = group1, Gene = gene, avg_log2FC,
         Pct_in_cluster = pct.1, Pct_out_cluster = pct.2,
         P_value = p_val, Adjusted_P_value = p_val_adj)
write.csv(fig1b_data, "Fig1B_TOP5_markers.csv", row.names = FALSE)
cat("Exported: Fig1B source data (", nrow(fig1b_data), "markers)\n")
```

---

# Figure 1C: Cell Composition by Thrombus Status

```{r fig1c-composition}
cat("\n--- Figure 1C: Cell composition by thrombus status (AD samples) ---\n")

# Subset to AD tissue only
ad_subset <- subset(seurat_obj, exp == 'Dissection tissue')

# Set factor levels
ad_subset$Presence_of_Thrombus <- factor(ad_subset$Presence_of_Thrombus,
                                         levels = c("No", "Yes"))

# Create stacked bar plot
p1c <- CellStatPlot(ad_subset,
                    stat.by = "subtype",
                    group.by = "Presence_of_Thrombus",
                    plot_type = "trend",
                    label = FALSE)

ggsave("Fig1C_composition_thrombus_AD.pdf", p1c, width = 8, height = 6)
cat("Saved: Fig1C_composition_thrombus_AD.pdf\n")

# Export source data
fig1c_data <- ad_subset@meta.data %>%
  group_by(Presence_of_Thrombus, subtype) %>%
  summarise(Cell_Count = n(), .groups = "drop") %>%
  group_by(Presence_of_Thrombus) %>%
  mutate(Percentage = Cell_Count / sum(Cell_Count) * 100)
write.csv(fig1c_data, "Fig1C_composition_by_thrombus.csv", row.names = FALSE)
cat("Exported: Fig1C source data\n")

print(p1c)
```

---

# Figure 1D: Volcano Plot of DEGs

```{r fig1d-volcano}
cat("\n--- Figure 1D: Volcano plot (thrombus Yes vs No in AD) ---\n")

# Run differential expression on AD samples
smc_ad_subset <- subset(ad_subset, subtype %in%
                          c('SMC1','SMC2','SMC3', 'SMC4','SMC5'))
smc_ad_subset$subtype <- droplevels(smc_ad_subset$subtype)

markers_smc_ad <- FindMarkers(
  smc_ad_subset,
  group.by = "Presence_of_Thrombus",
  ident.1 = "Yes", ident.2 = "No",
  test.use = "MAST",
  latent.vars = c("subtype")
)

# Select top genes to label
top_genes <- c("VEGFA","PLOD2","ANGPTL4","SPARC",
               "ELN","MYH10","ADAM33","MYH11")

p1d <- EnhancedVolcano(markers_smc_ad,
                       x = 'avg_log2FC',
                       y = 'p_val_adj',
                       lab = rownames(markers_smc_ad),
                       selectLab = top_genes,
                       xlim = c(-3, 3),
                       FCcutoff = 0.5,
                       legendPosition = 'right',
                       subtitle = NULL,
                       title = "Thrombus Yes vs No",
                       labSize = 4.0,
                       pointSize = 1.5,
                       drawConnectors = TRUE,
                       max.overlaps = Inf,
                       widthConnectors = 0.3)

ggsave("Fig1D_volcano_thrombus.pdf", p1d, width = 8, height = 6)
cat("Saved: Fig1D_volcano_thrombus.pdf\n")

# Export source data
fig1d_data <- markers_smc_ad %>%
  tibble::rownames_to_column("Gene") %>%
  select(Gene, avg_log2FC, Pct_Thrombus_Yes = pct.1, Pct_Thrombus_No = pct.2,
         P_value = p_val, Adjusted_P_value = p_val_adj) %>%
  arrange(Adjusted_P_value)
write.csv(fig1d_data, "Fig1D_DEG_thrombus_vs_no.csv", row.names = FALSE)
cat("Exported: Fig1D source data (", nrow(fig1d_data), "genes)\n")

print(p1d)
```

---

# Figure 1E: GSEA Analysis

```{r fig1e-gsea}
cat("\n--- Figure 1E: GSEA analysis ---\n")

smc_ad_subset <- RunDEtest(smc_ad_subset,
                           group.by = "Presence_of_Thrombus",
                           fc.threshold = 1,
                           test.use = 'MAST',
                           only.pos = FALSE)

smc_ad_subset <- RunGSEA(
  srt = smc_ad_subset,
  db = "GO_BP",
  group.by = "Presence_of_Thrombus",
  species = "Homo_sapiens",
  DE_threshold = "p_val_adj < 0.05"
)

p1e <- GSEAPlot(srt = smc_ad_subset, group.by = "Presence_of_Thrombus")

ggsave("Fig1E_GSEA_thrombus.pdf", p1e, width = 14, height = 8)
cat("Saved: Fig1E_GSEA_thrombus.pdf\n")

# Export source data
fig1e_data <- smc_ad_subset@tools[["GSEA_Presence_of_Thrombus_wilcox"]][["enrichment"]] %>%
  select(Group = Groups, Pathway_ID = ID, Description, setSize,
         enrichmentScore, NES, P_value = pvalue, Adjusted_P_value = p.adjust,
         core_enrichment) %>%
  group_by(Group) %>%
  filter(NES > 0) %>%
  arrange(Adjusted_P_value) %>%
  slice_head(n = 10) %>%
  ungroup()

write.csv(fig1e_data, "Fig1E_GSEA_TOP10_per_group.csv", row.names = FALSE)
cat("Exported: Fig1E source data\n")

print(p1e)
```

---

# Figure 1F: ANGPTL4 Expression

```{r fig1f-expression}
cat("\n--- Figure 1F: ANGPTL4 expression ---\n")

# Top panel: by thrombus status
p1f_top <- FeatureStatPlot(ad_subset,
                           stat.by = c("ANGPTL4"),
                           group.by = "subtype",
                           split.by = "Presence_of_Thrombus",
                           comparisons = TRUE,
                           stack = TRUE,
                           plot_type = "violin")

# Bottom panel: by tissue type (AD vs Adjacent)
p1f_bottom <- FeatureStatPlot(seurat_obj,
                              stat.by = c("ANGPTL4"),
                              group.by = "subtype",
                              split.by = "exp",
                              comparisons = TRUE,
                              stack = TRUE,
                              plot_type = "violin")

# Combine plots
p1f <- p1f_top / p1f_bottom

ggsave("Fig1F_ANGPTL4_violin.pdf", p1f, width = 12, height = 6)
cat("Saved: Fig1F_ANGPTL4_violin.pdf\n")

# Export source data with statistics
# Top panel
fig1f_top_stats <- ad_subset@meta.data %>%
  mutate(ANGPTL4 = FetchData(ad_subset, vars = "ANGPTL4")$ANGPTL4) %>%
  group_by(subtype) %>%
  wilcox_test(ANGPTL4 ~ Presence_of_Thrombus) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance() %>%
  select(Cell_Type = subtype, Comparison = .y.,
         Group1 = group1, Group2 = group2,
         n1, n2, statistic, P_value = p, Adjusted_P_value = p.adj, Significance = p.adj.signif)

# Bottom panel
fig1f_bottom_stats <- seurat_obj@meta.data %>%
  mutate(ANGPTL4 = FetchData(seurat_obj, vars = "ANGPTL4")$ANGPTL4) %>%
  group_by(subtype) %>%
  wilcox_test(ANGPTL4 ~ exp) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance() %>%
  select(Cell_Type = subtype, Comparison = .y.,
         Group1 = group1, Group2 = group2,
         n1, n2, statistic, P_value = p, Adjusted_P_value = p.adj, Significance = p.adj.signif)

# Combine both panels
fig1f_data <- bind_rows(
  fig1f_top_stats %>% mutate(Panel = "By_Thrombus_Status"),
  fig1f_bottom_stats %>% mutate(Panel = "By_Tissue_Type")
)

write.csv(fig1f_data, "Fig1F_ANGPTL4_statistics.csv", row.names = FALSE)
cat("Exported: Fig1F source data (", nrow(fig1f_data), "comparisons)\n")

print(p1f)
```

---

# Figure 1G: Myeloid UMAP

```{r fig1g-myeloid-umap}
cat("\n--- Figure 1G: Myeloid UMAP ---\n")

myeloid_obj$celltype <- factor(myeloid_obj$celltype,
                               levels = names(table(myeloid_obj$celltype)))

p1g <- CellDimPlot(myeloid_obj,
                   reduction = "umap",
                   group.by = "celltype",
                   label = TRUE,
                   pt.size = 0.5)

ggsave("Fig1G_myeloid_UMAP.pdf", p1g, width = 5.5, height = 4.5)
cat("Saved: Fig1G_myeloid_UMAP.pdf\n")

# Export source data
fig1g_data <- data.frame(
  UMAP_1 = myeloid_obj@reductions$umap@cell.embeddings[, 1],
  UMAP_2 = myeloid_obj@reductions$umap@cell.embeddings[, 2],
  Myeloid_Subtype = myeloid_obj$celltype
)
write.csv(fig1g_data, "Fig1G_myeloid_UMAP.csv", row.names = FALSE)
cat("Exported: Fig1G source data (", nrow(fig1g_data), "myeloid cells)\n")

print(p1g)
```

---

# Figure 1H: Myeloid Proportions

```{r fig1h-myeloid-proportions}
cat("\n--- Figure 1H: Myeloid proportions ---\n")

# Left: by thrombus status
myeloid_obj_sub <- subset(myeloid_obj, exp == 'Dissection tissue')

p1h_left <- CellStatPlot(myeloid_obj_sub,
                         stat.by = "celltype",
                         group.by = "Presence_of_Thrombus",
                         plot_type = "bar",
                         label = FALSE)

# Right: by tissue type
p1h_right <- CellStatPlot(myeloid_obj,
                          stat.by = "celltype",
                          group.by = "exp",
                          plot_type = "bar",
                          label = FALSE)

# Combine
p1h <- p1h_left | p1h_right

ggsave("Fig1H_myeloid_proportions.pdf", p1h, width = 8, height = 4)
cat("Saved: Fig1H_myeloid_proportions.pdf\n")

# Export source data
# By thrombus status
fig1h_thrombus <- myeloid_obj_sub@meta.data %>%
  group_by(Presence_of_Thrombus, celltype) %>%
  summarise(Cell_Count = n(), .groups = "drop") %>%
  group_by(Presence_of_Thrombus) %>%
  mutate(Percentage = Cell_Count / sum(Cell_Count) * 100)
write.csv(fig1h_thrombus, "Fig1H_myeloid_by_thrombus.csv", row.names = FALSE)

# By tissue type
fig1h_tissue <- myeloid_obj@meta.data %>%
  group_by(exp, celltype) %>%
  summarise(Cell_Count = n(), .groups = "drop") %>%
  group_by(exp) %>%
  mutate(Percentage = Cell_Count / sum(Cell_Count) * 100)
write.csv(fig1h_tissue, "Fig1H_myeloid_by_tissue.csv", row.names = FALSE)
cat("Exported: Fig1H source data\n")

print(p1h)
```

---

# Session Info

```{r session-info}
sessionInfo()
```

---

# Summary

All Figure 1 analyses completed successfully!

**Outputs:**
- 8 PDF figures (Fig1A-H)
- 10 CSV source data files
