---
title: "Figure 2: Identification and characterization of smooth muscle cell subpopulations in Aortic Dissection"
author: "Lin-Xiong Ye (445233812@qq.com)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)
```

# Overview

This notebook contains all analyses for **Figure 2**, which characterizes smooth muscle cell (SMC) subpopulations and their cell-cell communication patterns.

**Figure panels:**
- **2A**: Cell type distribution (AD vs Adjacent)
- **2B**: SMC subtype proportions
- **2C**: Hallmark pathway enrichment
- **2D**: C2 pathway comparison
- **2E**: Pseudotime trajectory heatmap
- **2F**: Cell-cell communication heatmap
- **2G**: SMC2/SMC3 signaling to endothelial cells
- **2H**: Information flow comparison

---

# Load Libraries

```{r load-libraries}
library(Seurat)
library(ggplot2)
library(dplyr)
library(SCP)
library(tidyverse)
library(CellChat)
library(ComplexHeatmap)
library(circlize)
library(patchwork)
library(monocle3)

options(future.globals.maxSize = 14000000 * 1024^2)
```

---

# Load Data

```{r load-data}
cat("=== Loading Data ===\n")

# Load preprocessed data
# Note: Update these paths to match your data location
seurat_obj <- readRDS('../seurat_object.rds')
result_cellchat <- readRDS('../cellchat_results.rds')

cat("Data loaded successfully!\n")
```

---


#-------------------------------------------------------------------------------
# Figure 2A: Cell type distribution (AD vs Adjacent)
#-------------------------------------------------------------------------------
cat("\n--- Figure 2A: Cell type distribution ---\n")

p2a <- CellStatPlot(seurat_obj,
                    stat.by = "subtype",
                    group.by = "exp",
                    plot_type = "trend",
                    label = FALSE)

ggsave("figures/figure2/Fig2A_celltype_distribution.pdf", p2a, width = 4, height = 4)
cat("Saved: Fig2A_celltype_distribution.pdf\n")

# Export source data for Fig2A
fig2a_data <- seurat_obj@meta.data %>%
  group_by(exp, subtype) %>%
  summarise(Cell_Count = n(), .groups = "drop") %>%
  group_by(exp) %>%
  mutate(Percentage = Cell_Count / sum(Cell_Count) * 100)
write.csv(fig2a_data, "figures/figure2/Fig2A_celltype_distribution.csv", row.names = FALSE)
cat("Exported: Fig2A source data\n")

#-------------------------------------------------------------------------------
# Figure 2B: SMC subtype proportions
#-------------------------------------------------------------------------------
cat("\n--- Figure 2B: SMC subtype proportions ---\n")

smc.mergeTEN = subset(seurat_obj, subtype %in% c('SMC1','SMC2','SMC3', 'SMC4','SMC5'))
smc.mergeTEN$subtype <- droplevels(smc.mergeTEN$subtype)

p2b <- CellStatPlot(smc.mergeTEN,
                    stat.by = "subtype",
                    group.by = "exp",
                    #plot_type = "trend",
                    label = FALSE)

ggsave("figures/figure2/Fig2B_SMC_proportions.pdf", p2b, width = 4, height = 4)
cat("Saved: Fig2B_SMC_proportions.pdf\n")

# Export source data for Fig2B
fig2b_data <- smc.mergeTEN@meta.data %>%
  group_by(exp, subtype) %>%
  summarise(Cell_Count = n(), .groups = "drop") %>%
  group_by(exp) %>%
  mutate(Percentage = Cell_Count / sum(Cell_Count) * 100)
write.csv(fig2b_data, "figures/figure2/Fig2B_SMC_proportions.csv", row.names = FALSE)
cat("Exported: Fig2B source data\n")

#-------------------------------------------------------------------------------
# Figure 2C-D: GSEA results (Hallmark and C2 pathways)
#-------------------------------------------------------------------------------
cat("\n--- Figure 2C-D: Pathway enrichment ---\n")

# Run enrichment analysis for SMC subtypes
smc.mergeTEN <- RunGSEA(
  srt = smc.mergeTEN, group.by = "subtype", 
  db = c("MSigDB_H",'MSigDB_C2'), species = "Homo_sapiens",
  DE_threshold = "p_val_adj < 0.05"
)

# Figure 2C: Hallmark pathways
p2c <- GSEAPlot(srt = smc.mergeTEN, group.by = "subtype",
                db = c("MSigDB_H"), plot_type = "comparison")

ggsave("figures/figure2/Fig2C_Hallmark_enrichment.pdf", p2c, width = 12, height = 10)
cat("Saved: Fig2C_Hallmark_enrichment.pdf\n")

# Export source data for Fig2C

# Filter Hallmark pathways and get TOP10 per SMC subtype
fig2c_data <- smc.mergeTEN@tools[["GSEA_subtype_wilcox"]][["enrichment"]] %>%
  filter(Database == "MSigDB_H") %>%
  select(SMC_Subtype = Groups, Pathway_ID = ID, Description, setSize,
         enrichmentScore, NES, P_value = pvalue, Adjusted_P_value = p.adjust,
         core_enrichment) %>%
  group_by(SMC_Subtype) %>%
  filter(NES > 0) %>%
  arrange(Adjusted_P_value) %>%
  slice_head(n = 10) %>%
  ungroup()

write.csv(fig2c_data, "figures/figure2/Fig2C_Hallmark_TOP10_per_subtype.csv", row.names = FALSE)
cat("Exported: Fig2C source data (TOP10 per subtype:", nrow(fig2c_data), "pathways)\n")


# Figure 2D: C2 pathways comparison
p2d <- GSEAPlot(srt = smc.mergeTEN, group.by = "subtype",
                db = c("MSigDB_C2"), plot_type = "comparison")

ggsave("figures/figure2/Fig2D_pathway_comparison.pdf", p2d, width = 12, height = 8)
cat("Saved: Fig2D_pathway_comparison.pdf\n")

# Export source data for Fig2D

# Filter C2 pathways and get TOP10 per SMC subtype
fig2d_data <- smc.mergeTEN@tools[["GSEA_subtype_wilcox"]][["enrichment"]] %>%
  filter(Database == "MSigDB_C2" & p.adjust < 0.05) %>%
  select(SMC_Subtype = Groups, Pathway_ID = ID, Description, setSize,
         enrichmentScore, NES, P_value = pvalue, Adjusted_P_value = p.adjust,
         core_enrichment) %>%
  group_by(SMC_Subtype) %>%
  filter(NES > 0) %>%
  arrange(Adjusted_P_value) %>%
  slice_head(n = 10) %>%
  ungroup()

write.csv(fig2d_data, "figures/figure2/Fig2D_C2_TOP10_per_subtype.csv", row.names = FALSE)
cat("Exported: Fig2D source data (TOP10 per subtype:", nrow(fig2d_data), "pathways)\n")


#-------------------------------------------------------------------------------
# Figure 2E: Pseudotime trajectory heatmap
#-------------------------------------------------------------------------------
cat("\n--- Figure 2E: Trajectory heatmap ---\n")

# Subsample cells for trajectory analysis (to reduce computation)
set.seed(123)
Idents(smc.mergeTEN) <- 'subtype'
cell_list <- Idents(smc.mergeTEN)

# Sample 1/5 of cells from each subtype
selectedCells <- sapply(split(Cells(smc.mergeTEN), cell_list), function(cells) {
  sample(cells, size = floor(length(cells) / 5))
})
selectedCells <- unlist(selectedCells)

# Create subsampled object
smc_sub <- subset(smc.mergeTEN, cells = selectedCells)

# Run Monocle3 trajectory analysis
smc_sub <- RunMonocle3(srt = smc_sub, clusters = "subtype")

# Extract trajectory components
trajectory <- smc_sub@tools$Monocle3$trajectory
milestones <- smc_sub@tools$Monocle3$milestones

# Visualize trajectory
p2e_trajectory <- CellDimPlot(smc_sub,
                              group.by = "Monocle3_partitions",
                              reduction = "UMAP",
                              label = TRUE,
                              theme_use = "theme_blank") +
  trajectory + milestones

# Pseudotime plot (this is actually Figure S2D)
p_s2d_pseudotime <- FeatureDimPlot(smc_sub,
                                   features = "Monocle3_Pseudotime",
                                   reduction = "UMAP",
                                   theme_use = "theme_blank") +
  trajectory

ggsave("figures/figureS2/FigS2D_pseudotime_UMAP.pdf", p_s2d_pseudotime, width = 5, height = 5)

# Annotate features for dynamic analysis
smc_sub <- AnnotateFeatures(smc_sub,
                             species = "Homo_sapiens",
                             db = c("TF"))

# Run dynamic feature analysis
smc_sub <- RunDynamicFeatures(srt = smc_sub,
                               lineages = c("Monocle3_Pseudotime"),
                               n_candidates = 100)

# Create dynamic heatmap (Figure 2E main panel)
Dynamic <- DynamicHeatmap(
  srt = smc_sub,
  lineages = c("Monocle3_Pseudotime"),
  use_fitted = TRUE,
  n_split = 1,
  species = "Homo_sapiens",
  db = c("GO_BP"),
  anno_terms = TRUE,
  anno_keys = TRUE,
  anno_features = TRUE,
  heatmap_palette = "viridis",
  cell_annotation = "subtype",
  separate_annotation = list("subtype"),
  separate_annotation_palette = c("Paired"),
  feature_annotation = c("TF"),
  feature_annotation_palcolor = list(c('red','#ca8687')),
  pseudotime_label = 25,
  pseudotime_label_color = "red",
  show_row_names = TRUE,
  row_names_side = "left",
  min_expcells = 10,
  r.sq = 0.05,
  dev.expl = 0.05,
  padjust = 0.05,
  height = 8.5,
  width = 2.5
)

ggsave("figures/figure2/Fig2E_trajectory_heatmap.pdf",
       Dynamic$plot,
       width = 25,
       height = 15)

cat("Saved: Fig2E trajectory analysis\n")

# Export source data for Fig2E
# Cell pseudotime information
fig2e_pseudotime <- data.frame(
  Pseudotime = smc_sub$Monocle3_Pseudotime,
  SMC_Subtype = smc_sub$subtype
)
write.csv(fig2e_pseudotime, "figures/figure2/Fig2E_cell_pseudotime.csv", row.names = FALSE)


#-------------------------------------------------------------------------------
# Figure 2F: Cell-cell communication heatmap (interactions_heatmap)
#-------------------------------------------------------------------------------
cat("\n--- Figure 2F: Communication heatmap ---\n")

# Merge CellChat objects (control first, test second)
object.list <- list(Normal = result_cellchat$cellchat_control,
                    AD = result_cellchat$cellchat_test)
cellchat <- mergeCellChat(object.list, add.names = names(object.list))

cat("Merged CellChat object created\n")

# Create interaction heatmaps
gg1 <- netVisual_heatmap(cellchat)
gg2 <- netVisual_heatmap(cellchat, measure = "weight")

# Save combined heatmap (Figure 2F)
pdf("figures/figure2/Fig2F_interactions_heatmap.pdf", width = 6, height = 3.7)
draw(gg1 + gg2, ht_gap = unit(1, "cm"))
dev.off()

# Export source data for Fig2F
cellchat_control <- result_cellchat$cellchat_control
cellchat_test <- result_cellchat$cellchat_test

# Interaction counts (Number of interactions)
fig2f_count_normal <- as.data.frame(cellchat_control@net$count) %>%
  tibble::rownames_to_column("Source") %>%
  pivot_longer(-Source, names_to = "Target", values_to = "Interaction_Count") %>%
  mutate(Condition = "Normal")

fig2f_count_ad <- as.data.frame(cellchat_test@net$count) %>%
  tibble::rownames_to_column("Source") %>%
  pivot_longer(-Source, names_to = "Target", values_to = "Interaction_Count") %>%
  mutate(Condition = "AD")

fig2f_count_data <- bind_rows(fig2f_count_normal, fig2f_count_ad)
write.csv(fig2f_count_data, "figures/figure2/Fig2F_interaction_counts.csv", row.names = FALSE)

# Interaction weights (Strength of interactions)
fig2f_weight_normal <- as.data.frame(cellchat_control@net$weight) %>%
  tibble::rownames_to_column("Source") %>%
  pivot_longer(-Source, names_to = "Target", values_to = "Interaction_Weight") %>%
  mutate(Condition = "Normal")

fig2f_weight_ad <- as.data.frame(cellchat_test@net$weight) %>%
  tibble::rownames_to_column("Source") %>%
  pivot_longer(-Source, names_to = "Target", values_to = "Interaction_Weight") %>%
  mutate(Condition = "AD")

fig2f_weight_data <- bind_rows(fig2f_weight_normal, fig2f_weight_ad)
write.csv(fig2f_weight_data, "figures/figure2/Fig2F_interaction_weights.csv", row.names = FALSE)

cat("Exported: Fig2F source data (counts and weights)\n")

#-------------------------------------------------------------------------------
# Figure 2G: SMC2/SMC3 signaling to endothelial cells
#-------------------------------------------------------------------------------
cat("\n--- Figure 2G: SMC to EC signaling ---\n")


cellchat_ad <- result_cellchat$cellchat_test

# Bubble plot showing SMC2/SMC3 to EC signaling
p2g <- netVisual_bubble(cellchat_ad,
                        sources.use = c(6, 7),  # SMC2 and SMC3
                        targets.use = 1,         # Endothelial cells
                        remove.isolate = FALSE)

ggsave("figures/figure2/Fig2G_SMC_EC_signaling.pdf", p2g, width = 3, height = 4)

cat("Saved: Fig2G_SMC_EC_signaling.pdf\n")

# Export source data for Fig2G
# Extract ligand-receptor pairs from SMC2/SMC3 to EC with probability and p-values

# Extract all pathways
pathway_names <- dimnames(cellchat_ad@net$prob)[[3]]

# Create a list to store results
interaction_list <- list()

for (pathway in pathway_names) {
  # Get probability and p-value matrices for this pathway
  prob_matrix <- cellchat_ad@net$prob[, , pathway]
  pval_matrix <- cellchat_ad@net$pval[, , pathway]

  # Extract SMC2 to Endothelial cells interactions
  if ("SMC2" %in% rownames(prob_matrix) && "Endothelial cells" %in% colnames(prob_matrix)) {
    smc2_ec_prob <- prob_matrix["SMC2", "Endothelial cells"]
    smc2_ec_pval <- pval_matrix["SMC2", "Endothelial cells"]

    if (smc2_ec_prob > 0) {
      interaction_list[[length(interaction_list) + 1]] <- data.frame(
        Source = "SMC2",
        Target = "Endothelial cells",
        Pathway = pathway,
        Probability = smc2_ec_prob,
        P_value = smc2_ec_pval
      )
    }
  }

  # Extract SMC3 to Endothelial cells interactions
  if ("SMC3" %in% rownames(prob_matrix) && "Endothelial cells" %in% colnames(prob_matrix)) {
    smc3_ec_prob <- prob_matrix["SMC3", "Endothelial cells"]
    smc3_ec_pval <- pval_matrix["SMC3", "Endothelial cells"]

    if (smc3_ec_prob > 0) {
      interaction_list[[length(interaction_list) + 1]] <- data.frame(
        Source = "SMC3",
        Target = "Endothelial cells",
        Pathway = pathway,
        Probability = smc3_ec_prob,
        P_value = smc3_ec_pval
      )
    }
  }
}

# Combine all interactions
if (length(interaction_list) > 0) {
  fig2g_data <- bind_rows(interaction_list) %>%
    arrange(desc(Probability))

  write.csv(fig2g_data, "figures/figure2/Fig2G_SMC_EC_signaling.csv", row.names = FALSE)
  cat("Exported: Fig2G source data (", nrow(fig2g_data), "interactions)\n")
} else {
  cat("Warning: No SMC2/SMC3 to EC interactions found\n")
}



#-------------------------------------------------------------------------------
# Figure 2H: information flow comparison
#-------------------------------------------------------------------------------
cat("\n--- Figure 2H: VEGFA signaling shift ---\n")
# Rank net for VEGF signaling comparison
# This shows the shift in VEGF signaling sources
p1 <- netVisual_bubble(cellchat, 
                       sources.use = c(2,3,4,5,6,7,8,9),  targets.use = c(1), 
                       comparison = c(1, 2), max.dataset = 1, remove.isolate = T,
                       title.name = paste("Increased signaling in ",
                                          names(object.list)[1]), angle.x = 45)
p2 <- netVisual_bubble(cellchat, 
                       sources.use = c(2,3,4,5,6,7,8,9),  targets.use = c(1), 
                       comparison = c(1, 2), max.dataset = 2, 
                       title.name = paste("Increased signaling in ",
                                          names(object.list)[2]), angle.x = 45, remove.isolate = T)
p2h <- p1 + p2

ggsave("figures/figure2/Fig2H_information_flow.pdf", p2h, width = 9, height = 4)

cat("Saved: Fig2H_VEGFA_signaling_shift.pdf\n")

# Export source data for Fig2H
# Extract ALL pathways to Endothelial cells with probability and p-values
if (!is.null(cellchat_control@net$prob) && !is.null(cellchat_control@net$pval) &&
    !is.null(cellchat_test@net$prob) && !is.null(cellchat_test@net$pval)) {

  # Get pathway names from both conditions
  pathway_names_normal <- dimnames(cellchat_control@net$prob)[[3]]
  pathway_names_ad <- dimnames(cellchat_test@net$prob)[[3]]

  # Get union of all pathways
  all_pathways <- unique(c(pathway_names_normal, pathway_names_ad))

  # Create lists to store results
  interaction_list_normal <- list()
  interaction_list_ad <- list()

  for (pathway in all_pathways) {
    # Normal condition - check if pathway exists
    if (pathway %in% pathway_names_normal) {
      prob_matrix_normal <- cellchat_control@net$prob[, , pathway]
      pval_matrix_normal <- cellchat_control@net$pval[, , pathway]

      # Extract all cell types to Endothelial cells (Normal)
      if ("Endothelial cells" %in% colnames(prob_matrix_normal)) {
        for (source_cell in rownames(prob_matrix_normal)) {
          prob_val <- prob_matrix_normal[source_cell, "Endothelial cells"]
          pval_val <- pval_matrix_normal[source_cell, "Endothelial cells"]

          if (prob_val > 0) {
            interaction_list_normal[[length(interaction_list_normal) + 1]] <- data.frame(
              Source = source_cell,
              Target = "Endothelial cells",
              Pathway = pathway,
              Probability = prob_val,
              P_value = pval_val,
              Condition = "Normal"
            )
          }
        }
      }
    }

    # AD condition - check if pathway exists
    if (pathway %in% pathway_names_ad) {
      prob_matrix_ad <- cellchat_test@net$prob[, , pathway]
      pval_matrix_ad <- cellchat_test@net$pval[, , pathway]

      # Extract all cell types to Endothelial cells (AD)
      if ("Endothelial cells" %in% colnames(prob_matrix_ad)) {
        for (source_cell in rownames(prob_matrix_ad)) {
          prob_val <- prob_matrix_ad[source_cell, "Endothelial cells"]
          pval_val <- pval_matrix_ad[source_cell, "Endothelial cells"]

          if (prob_val > 0) {
            interaction_list_ad[[length(interaction_list_ad) + 1]] <- data.frame(
              Source = source_cell,
              Target = "Endothelial cells",
              Pathway = pathway,
              Probability = prob_val,
              P_value = pval_val,
              Condition = "AD"
            )
          }
        }
      }
    }
  }

  # Combine both conditions
  if (length(interaction_list_normal) > 0 || length(interaction_list_ad) > 0) {
    fig2h_data <- bind_rows(
      if (length(interaction_list_normal) > 0) bind_rows(interaction_list_normal) else NULL,
      if (length(interaction_list_ad) > 0) bind_rows(interaction_list_ad) else NULL
    ) %>%
      arrange(Condition, desc(Probability))

    write.csv(fig2h_data, "figures/figure2/Fig2H_all_pathways_to_EC.csv", row.names = FALSE)
    cat("Exported: Fig2H source data (", nrow(fig2h_data), "interactions across all pathways)\n")
  } else {
    cat("Warning: No interactions to Endothelial cells found\n")
  }
} else {
  cat("Warning: Fig2H probability or p-value data not available\n")
}



#===============================================================================

---

# Session Info

```{r session-info}
sessionInfo()
```
